The CrypTool Script: \\ Cryptography, Mathematics and more \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free eLearning program CrypTool\\ (9th edition -- distributed with CrypTool version 1.4.20)} \vspace{60pt}

### 2009-06-30 script/en/chapters/numbertheory.tex
Use the book class for the script, instead of the article class

\newpage
\hypertarget{Chapter_ElementaryNT}{}
\section{Introduction to Elementary Number Theory with Examples}
\label{Chapter_ElementaryNT}
(Bernhard Esslinger, July 2001; Updates: Dec. 2001, June 2002, May 2003,
 May 2005, March 2006, June 2007) \\

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Mathematics and cryptography}
A large proportion of modern, asymmetric cryptography\index{Encryption!asymmetric} is based on mathematical knowledge -- on the properties 
(``laws'') of whole numbers, which are investigated in elementary \index{Number theory!elementary} number 
theory. Here, the word ``elementary'' means that questions raised in number theory are essentially rooted in the set 
of natural and whole numbers.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Introduction to number theory} \index{Number theory!introduction}

@@

% --------------------------------------------------------------------------
\newpage
\subsubsection{Convention}
Unless stated otherwise: 
\begin{itemize}
\item The letters $a, b, c, d, e, k, n, m, p, q$ are used to present whole numbers.
\item The letters $i$ ~\mbox{and} $j$ represent natural numbers. 
\item The letters $p$ always represents a prime number.
\item The sets $\mathbb{N} = \{ 1, 2, 3, \cdots \}$ and $\mathbb{Z} =\{ \cdots, -3, -2, -1, 0, 1, 2, 3, \cdots \}$ 
are the {\em natural} and {\em whole} numbers respectively.
\end{itemize}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Prime numbers and the first fundamental theorem of elementary number theory}
\index{Number theory!elementary} Many of the problems in elementary number theory are concerned with prime 
numbers.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection[Divisibility, modulus and remainder classes]
           {Divisibility, modulus and remainder classes\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can have a
    playful view at the calculation with congruences, discussed in this and
    the next chapter (see learning unit 2.1, pages 2-9/40).\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.
}
\index{Modulus} \index{Divisibility}

@@

% --------------------------------------------------------------------------
\subsubsection{The modulo operation -- working with congruence} \index{Congruence}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Calculations with finite sets}

% --------------------------------------------------------------------------
\subsubsection{Laws of modular calculations}

@@

% --------------------------------------------------------------------------
\hypertarget{Chapter_ElementaryNT_5_2}{}
\subsubsection{Patterns and structures}
\label{Label_Chapter_ElementaryNT_5_2}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
% \pagebreak
\subsection{Examples of modular calculations}

@@

% --------------------------------------------------------------------------
\subsubsection{Addition and multiplication} \index{Addition} \index{Multiplication}
\label{addmult}

@@

% --------------------------------------------------------------------------
\subsubsection*{Example of an addition table}
The result when we add $3$ and $4 {\rm ~(mod~ } 5)$ is calculated as follows: Calculate $3 + 4 = 7$ and keep 
subtracting $5$ from the result until the result is less than the modulo:  $7 - 5 = 2$. 
Therefore:  $3 + 4 \equiv 2 {\rm ~(mod~ } 5)$.
%
\begin{table}[ht]
\begin{center}
\begin{tabular}{r|ccccc}
Addition table modulo 5: \quad + &  0 & 1 & 2 & 3 & 4  \\
\hline
0 &  0 & 1 & 2 & 3 & 4 \\  
1 & 1 &  2 & 3 & 4 & 0 \\
2 & 2 & 3 & 4 & 0 & 1 \\
3 & 3 & 4 & 0 & 1 & 2 \\
4 & 4 & 0 & 1 & 2 & 3 \\[-5ex] % remove some extra space 
\end{tabular} 
\end{center} 
\end{table}

% --------------------------------------------------------------------------
\subsubsection*{Example of a multiplication table:}
The result of the multiplication $4 * 4 {\rm ~(mod~ } 5)$ is calculated as follows: $4 * 4 = 
16$ and subtract $5$ until the result is less than the modulus.
$$16 - 5 = 11;~ 11 - 5 = 6;~6- 5 = 1.$$
The table directly shows that $4 * 4 \equiv 1 {\rm ~(mod~} 5)$ because $16 : 5 = 3$ remainder $1$. 
Multiplication is defined on the set $\mathbb{Z}$  excluding $0$.
%
\begin{table}[ht]
\begin{center}
\begin{tabular}{r|cccc}
Multiplication table modulo 5: \quad * & 1& 2 & 3 & 4  \\
\hline 
1 & 1 &    2    &    3    & 4 \\
2 & 2 & {\bf 4} & {\bf 1} & 3 \\ 
3 & 3 & {\bf 1} & {\bf 4} & 2  \\
4 & 4 &    3    &    2    & 1 
\end{tabular}
\end{center} 
\end{table}

% --------------------------------------------------------------------------
\subsubsection{Additive and multiplicative inverses}

@@

\subsubsection*{Table 1: Multiplication table modulo $17$ (for $a=5$ and $a=6$)}\label{SrcArith1a}
\begin{center}
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                   & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17 & 18 \\
\hline
\hline  
$5*i$                & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80  & 85 & 90   \\
remainder                & 5 & 10 & 15  & 3  & 8 & 13  & {\bf 1}  & 6 & 11 & 16  & 4  & 9 & 14  & 2  & 7 & 12 & 0  & 5   \\
\hline
$6*i$                 & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108   \\
remainder                & 6 & 12  & {\bf 1}  & 7 & 13  & 2  & 8 & 14  & 3  & 9 & 15  & 4 & 10 & 16  & 5 & 11 & 0  & 6   \\
\hline
\end{tabular}
\end{center} 
Between $i=1, \cdots, m$, all values between $0, \cdots, m-1$ occur for the remainders, because both $5$ and $6$ are 
also relatively prime\index{Relatively prime} to the modulus $m=17$.
\enlargethispage{0.5cm}

@@

\subsubsection*{Table 2: Multiplication table modulo $13$ (for $a=5$ and $a=6$)}\label{SrcArith1b}
\begin{center}                                                                          
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                    & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17  & 18 \\
\hline 
\hline 
$5*i$                  & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80 & 85  & 90 \\
remainder                 & 5 & 10  & 2  & 7  & 12  & 4 & 9  & {\bf 1}  & 6  & 11 & 3  & 8  & 0 & 5  & 10  & 2   & 7   & 12 \\
\hline 
$6*i$                  & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108 \\
remainder                 & 6  & 12  & 5  & 11  & 4  & 10  & 3  & 9  & 2  & 8  & {\bf 1}  & 7  & 0  & 6  & 12  & 5   & 11   & 4 \\
\hline 
\end{tabular}
\end{center}

@@

\subsubsection*{Table 3: Multiplication table modulo $12$ (for $a=5$ and $a=6$)}
\begin{center}                                                                          
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                    & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17 & 18 \\
\hline 
\hline 
$5*i$                  & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80 & 85  & 90 \\
remainder                 & 5 & 10  & 3  & 8  & {\bf 1}  & 6 & 11  & 4  & 9  & 2  & 7  & 0  & 5 & 10  & 3 & 8   & 1   & 6 \\
\hline 
$6*i$                  & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108 \\
remainder                 & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0   & 6   & 0 \\
\hline 
\end{tabular}
\end{center} 
We have calculated $(5 * i)$ (mod $12$)  and  $(6*i)$ (mod $12$).
Between $i=1, \cdots, m$, not all values between $0, \cdots, m-1$ occur and $6$ does not have an inverse mod $12$, 
because $6$ and the modulus $m=12$ are not co-prime\index{Number!co-prime}.

@@

% --------------------------------------------------------------------------
\subsubsection{Raising to the power} \index{Raising to the power} 
In modular arithmetic, raising to the power is defined as repeated multiplication -- as usual except that 
multiplication is now slightly different.  We can even apply the usual rules, such as:
\begin{eqnarray*}
a^{b+c} & = & a^b * a^c,  \nonumber \\
(a^b)^c & = & a^{b*c} = a^{c*b} = (a^c)^b. \nonumber
\end{eqnarray*}

@@

% --------------------------------------------------------------------------
\vskip +10pt
\hypertarget{hohpot}{}
\subsubsection{Fast calculation of high powers} \label{hohpot}\index{Power}
%\subsubsection*{3.6.3.1 ~~~Fast calculation of high powers} \hypertarget{hohpot}{} \index{Power}
%\addcontentsline{toc}{subsubsection}{~~~~~~~~~~3.6.3.1 ~Fast calculation of high powers}
RSA encryption and decryption\footnote{See chapter \ref{rsabeweis} (Proof of the
RSA procedure with Euler-Fermat) \index{RSA} and chapter~\ref{rsaconcrete} (The
RSA procedure with actual numbers).} entails calculating high powers modulo
$m$.  For example, the calculation ($100^5) {\rm ~(mod~} 3)$ exceeds the 32-bit
long integer\index{Long integer} number range provided we calculate $a^n$ by
actually multiplying a with itself $n$ times in line with the definition. In
the case of extremely large numbers, even a fast computer chip would take
longer than the age of the universe to calculate a single exponential. Luckily,
there is an extremely effective shortcut for calculating exponentials (but not
for calculating logarithms).

@@

% --------------------------------------------------------------------------
\subsubsection{Roots and logarithms} \index{Root} \index{Logarithm}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Groups and modular arithmetic in $\mathbb{Z}_n$ and $\mathbb{Z}_n^*$}
\index{Group}
Mathematical ``{\em groups}'' play a decisive role in number theory and cryptography. We only talk of groups if, for a 
defined set and a defined relation (an operation such as addition or multiplication), the following properties are fulfilled:

@@

% --------------------------------------------------------------------------
\subsubsection{Addition in a group}\index{Addition}

@@

% --------------------------------------------------------------------------
\subsubsection{Multiplication in a group}\index{Multiplication}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{Chapter_ElementaryNT_8}{}
\subsection{Euler function, Fermat's little theorem and Euler-Fermat}

% --------------------------------------------------------------------------
\hypertarget{patternsandstructures}{}
\subsubsection{Patterns and structures} \index{Structure}
\label{patternsandstructures}
As mathematicians investigate the structure $a *x \equiv b$ mod $m$
(see \hyperlink{Chapter_ElementaryNT_5_2}
{chapter \ref{Label_Chapter_ElementaryNT_5_2}}), so they are interested in
the structure $ x^{a} \equiv b$ mod $m.$

@@

% --------------------------------------------------------------------------
\subsubsection{The Euler function\index{Euler!(phi) function}}
\label{L-Euler-Function}
Given $n$, the number of numbers from the set $\{1, \cdots, n-1\}$ that are relatively prime to $n$ is equal to the value of 
the Euler\footnote{Leonhard Euler, Swiss mathematician, Apr 15, 1707 -- Sep 18, 1783\index{Euler, Leonhard}
} function $J(n)$.

@@

% --------------------------------------------------------------------------
\subsubsection{The theorem of Euler-Fermat}\index{Euler, Leonhard}\index{RSA}
\label{Label_KleinerSatzFermat-chap3}
In order to prove the RSA procedure, we need Fermat's theorem and its 
generalisation (Euler-Fermat theorem) -- 
\hyperlink{KleinerSatzFermat-chap2}{please see chapter \ref{primality_tests}}.

@@

% --------------------------------------------------------------------------
\subsubsection{Calculation of the multiplicative inverse}

@@

% --------------------------------------------------------------------------
\subsubsection{Fixpoints\index{Fixpoint} modulo 26}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
%\pagebreak
\hypertarget{Chapter_ElementaryNT_9}{}
\subsection[Multiplicative order and primitive roots]
           {Multiplicative order and primitive roots\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can have a
    playful experience with primitive roots
    (see learning unit 2.2, pages 10-14/40 and 24-40/40).\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.
}
\label{MultOrdPrimitveRoot}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{Chapter_ElementaryNT_10}{}
\hypertarget{RSABeweis}{}
\subsection{Proof of the RSA procedure with Euler-Fermat}
\index{RSA}\label{rsabeweis}
Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA!RSA procedure} is the most
common asymmetric cryptography procedure. Developed in 1978 by Ronald Rivest,
Adi Shamir and Leonard Adleman, it can be used both for signatures and for
encryption. Cryptographers always associate this procedure with the
abbreviation ``\textbf{RSA}'' $-$ the following remark is meant with humor to show
that each letter combination can be used with several meanings: In Britain
the ``Royal Society for the encouragement of Arts, Manufactures \& Commerce''
is commonly known as the ``RSA''.
}$^,$\footnote{%
In literature\index{Literature} and in movies\index{Movies} not only 
classic but also modern cryptographic methods have been used 
%(\hyperlink{appendix-movies}
%{see appendix A3: Movies and Literature with Relation to Cryptography}).
(see appendix \ref{s:appendix-movies}).
% hier die items nicht einrücken ! xxxxxxxxxxxxxxx
% \begin{list}{\textbullet}{\leftmargin10pt}
%   \item the movie ``House of Cards'', 1992
%   \item the story ``The Dialogue of the Sisters'' by Dr.~C.~Elsner
% \end{list}\vspace{-\baselineskip}\vspace{-\parskip}
} %end footnote
procedure in the group $\mathbb{Z}_n^*$.

>>>>

\newpage
\hypertarget{Chapter_ElementaryNT}{}
\chapter{Introduction to Elementary Number Theory with Examples}
\label{Chapter_ElementaryNT}
(Bernhard Esslinger, July 2001; Updates: Dec. 2001, June 2002, May 2003,
 May 2005, March 2006, June 2007) \\

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{Mathematics and cryptography}
A large proportion of modern, asymmetric cryptography\index{Encryption!asymmetric} is based on mathematical knowledge -- on the properties 
(``laws'') of whole numbers, which are investigated in elementary \index{Number theory!elementary} number 
theory. Here, the word ``elementary'' means that questions raised in number theory are essentially rooted in the set 
of natural and whole numbers.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{Introduction to number theory} \index{Number theory!introduction}

@@

% --------------------------------------------------------------------------
\newpage
\subsection{Convention}
Unless stated otherwise: 
\begin{itemize}
\item The letters $a, b, c, d, e, k, n, m, p, q$ are used to present whole numbers.
\item The letters $i$ ~\mbox{and} $j$ represent natural numbers. 
\item The letters $p$ always represents a prime number.
\item The sets $\mathbb{N} = \{ 1, 2, 3, \cdots \}$ and $\mathbb{Z} =\{ \cdots, -3, -2, -1, 0, 1, 2, 3, \cdots \}$ 
are the {\em natural} and {\em whole} numbers respectively.
\end{itemize}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{Prime numbers and the first fundamental theorem of elementary number theory}
\index{Number theory!elementary} Many of the problems in elementary number theory are concerned with prime 
numbers.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section[Divisibility, modulus and remainder classes]
           {Divisibility, modulus and remainder classes\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can have a
    playful view at the calculation with congruences, discussed in this and
    the next chapter (see learning unit 2.1, pages 2-9/40).\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.
}
\index{Modulus} \index{Divisibility}

@@

% --------------------------------------------------------------------------
\subsection{The modulo operation -- working with congruence} \index{Congruence}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{Calculations with finite sets}

% --------------------------------------------------------------------------
\subsection{Laws of modular calculations}

@@

% --------------------------------------------------------------------------
\hypertarget{Chapter_ElementaryNT_5_2}{}
\subsection{Patterns and structures}
\label{Label_Chapter_ElementaryNT_5_2}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
% \pagebreak
\section{Examples of modular calculations}

@@

% --------------------------------------------------------------------------
\subsection{Addition and multiplication} \index{Addition} \index{Multiplication}
\label{addmult}

@@

% --------------------------------------------------------------------------
\subsection*{Example of an addition table}
The result when we add $3$ and $4 {\rm ~(mod~ } 5)$ is calculated as follows: Calculate $3 + 4 = 7$ and keep 
subtracting $5$ from the result until the result is less than the modulo:  $7 - 5 = 2$. 
Therefore:  $3 + 4 \equiv 2 {\rm ~(mod~ } 5)$.
%
\begin{table}[ht]
\begin{center}
\begin{tabular}{r|ccccc}
Addition table modulo 5: \quad + &  0 & 1 & 2 & 3 & 4  \\
\hline
0 &  0 & 1 & 2 & 3 & 4 \\  
1 & 1 &  2 & 3 & 4 & 0 \\
2 & 2 & 3 & 4 & 0 & 1 \\
3 & 3 & 4 & 0 & 1 & 2 \\
4 & 4 & 0 & 1 & 2 & 3 \\[-5ex] % remove some extra space 
\end{tabular} 
\end{center} 
\end{table}

% --------------------------------------------------------------------------
\subsection*{Example of a multiplication table:}
The result of the multiplication $4 * 4 {\rm ~(mod~ } 5)$ is calculated as follows: $4 * 4 = 
16$ and subtract $5$ until the result is less than the modulus.
$$16 - 5 = 11;~ 11 - 5 = 6;~6- 5 = 1.$$
The table directly shows that $4 * 4 \equiv 1 {\rm ~(mod~} 5)$ because $16 : 5 = 3$ remainder $1$. 
Multiplication is defined on the set $\mathbb{Z}$  excluding $0$.
%
\begin{table}[ht]
\begin{center}
\begin{tabular}{r|cccc}
Multiplication table modulo 5: \quad * & 1& 2 & 3 & 4  \\
\hline 
1 & 1 &    2    &    3    & 4 \\
2 & 2 & {\bf 4} & {\bf 1} & 3 \\ 
3 & 3 & {\bf 1} & {\bf 4} & 2  \\
4 & 4 &    3    &    2    & 1 
\end{tabular}
\end{center} 
\end{table}

% --------------------------------------------------------------------------
\subsection{Additive and multiplicative inverses}

@@

\subsection*{Table 1: Multiplication table modulo $17$ (for $a=5$ and $a=6$)}\label{SrcArith1a}
\begin{center}
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                   & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17 & 18 \\
\hline
\hline  
$5*i$                & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80  & 85 & 90   \\
remainder                & 5 & 10 & 15  & 3  & 8 & 13  & {\bf 1}  & 6 & 11 & 16  & 4  & 9 & 14  & 2  & 7 & 12 & 0  & 5   \\
\hline
$6*i$                 & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108   \\
remainder                & 6 & 12  & {\bf 1}  & 7 & 13  & 2  & 8 & 14  & 3  & 9 & 15  & 4 & 10 & 16  & 5 & 11 & 0  & 6   \\
\hline
\end{tabular}
\end{center} 
Between $i=1, \cdots, m$, all values between $0, \cdots, m-1$ occur for the remainders, because both $5$ and $6$ are 
also relatively prime\index{Relatively prime} to the modulus $m=17$.
\enlargethispage{0.5cm}

@@

\subsection*{Table 2: Multiplication table modulo $13$ (for $a=5$ and $a=6$)}\label{SrcArith1b}
\begin{center}                                                                          
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                    & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17  & 18 \\
\hline 
\hline 
$5*i$                  & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80 & 85  & 90 \\
remainder                 & 5 & 10  & 2  & 7  & 12  & 4 & 9  & {\bf 1}  & 6  & 11 & 3  & 8  & 0 & 5  & 10  & 2   & 7   & 12 \\
\hline 
$6*i$                  & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108 \\
remainder                 & 6  & 12  & 5  & 11  & 4  & 10  & 3  & 9  & 2  & 8  & {\bf 1}  & 7  & 0  & 6  & 12  & 5   & 11   & 4 \\
\hline 
\end{tabular}
\end{center}

@@

\subsection*{Table 3: Multiplication table modulo $12$ (for $a=5$ and $a=6$)}
\begin{center}                                                                          
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                    & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17 & 18 \\
\hline 
\hline 
$5*i$                  & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80 & 85  & 90 \\
remainder                 & 5 & 10  & 3  & 8  & {\bf 1}  & 6 & 11  & 4  & 9  & 2  & 7  & 0  & 5 & 10  & 3 & 8   & 1   & 6 \\
\hline 
$6*i$                  & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108 \\
remainder                 & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0   & 6   & 0 \\
\hline 
\end{tabular}
\end{center} 
We have calculated $(5 * i)$ (mod $12$)  and  $(6*i)$ (mod $12$).
Between $i=1, \cdots, m$, not all values between $0, \cdots, m-1$ occur and $6$ does not have an inverse mod $12$, 
because $6$ and the modulus $m=12$ are not co-prime\index{Number!co-prime}.

@@

% --------------------------------------------------------------------------
\subsection{Raising to the power} \index{Raising to the power} 
In modular arithmetic, raising to the power is defined as repeated multiplication -- as usual except that 
multiplication is now slightly different.  We can even apply the usual rules, such as:
\begin{eqnarray*}
a^{b+c} & = & a^b * a^c,  \nonumber \\
(a^b)^c & = & a^{b*c} = a^{c*b} = (a^c)^b. \nonumber
\end{eqnarray*}

@@

% --------------------------------------------------------------------------
\vskip +10pt
\hypertarget{hohpot}{}
\subsection{Fast calculation of high powers} \label{hohpot}\index{Power}
%\subsubsection*{3.6.3.1 ~~~Fast calculation of high powers} \hypertarget{hohpot}{} \index{Power}
%\addcontentsline{toc}{subsubsection}{~~~~~~~~~~3.6.3.1 ~Fast calculation of high powers}
RSA encryption and decryption\footnote{See chapter \ref{rsabeweis} (Proof of the
RSA procedure with Euler-Fermat) \index{RSA} and chapter~\ref{rsaconcrete} (The
RSA procedure with actual numbers).} entails calculating high powers modulo
$m$.  For example, the calculation ($100^5) {\rm ~(mod~} 3)$ exceeds the 32-bit
long integer\index{Long integer} number range provided we calculate $a^n$ by
actually multiplying a with itself $n$ times in line with the definition. In
the case of extremely large numbers, even a fast computer chip would take
longer than the age of the universe to calculate a single exponential. Luckily,
there is an extremely effective shortcut for calculating exponentials (but not
for calculating logarithms).

@@

% --------------------------------------------------------------------------
\subsection{Roots and logarithms} \index{Root} \index{Logarithm}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{Groups and modular arithmetic in $\mathbb{Z}_n$ and $\mathbb{Z}_n^*$}
\index{Group}
Mathematical ``{\em groups}'' play a decisive role in number theory and cryptography. We only talk of groups if, for a 
defined set and a defined relation (an operation such as addition or multiplication), the following properties are fulfilled:

@@

% --------------------------------------------------------------------------
\subsection{Addition in a group}\index{Addition}

@@

% --------------------------------------------------------------------------
\subsection{Multiplication in a group}\index{Multiplication}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{Chapter_ElementaryNT_8}{}
\section{Euler function, Fermat's little theorem and Euler-Fermat}

% --------------------------------------------------------------------------
\hypertarget{patternsandstructures}{}
\subsection{Patterns and structures} \index{Structure}
\label{patternsandstructures}
As mathematicians investigate the structure $a *x \equiv b$ mod $m$
(see \hyperlink{Chapter_ElementaryNT_5_2}
{chapter \ref{Label_Chapter_ElementaryNT_5_2}}), so they are interested in
the structure $ x^{a} \equiv b$ mod $m.$

@@

% --------------------------------------------------------------------------
\subsection{The Euler function\index{Euler!(phi) function}}
\label{L-Euler-Function}
Given $n$, the number of numbers from the set $\{1, \cdots, n-1\}$ that are relatively prime to $n$ is equal to the value of 
the Euler\footnote{Leonhard Euler, Swiss mathematician, Apr 15, 1707 -- Sep 18, 1783\index{Euler, Leonhard}
} function $J(n)$.

@@

% --------------------------------------------------------------------------
\subsection{The theorem of Euler-Fermat}\index{Euler, Leonhard}\index{RSA}
\label{Label_KleinerSatzFermat-chap3}
In order to prove the RSA procedure, we need Fermat's theorem and its 
generalisation (Euler-Fermat theorem) -- 
\hyperlink{KleinerSatzFermat-chap2}{please see chapter \ref{primality_tests}}.

@@

% --------------------------------------------------------------------------
\subsection{Calculation of the multiplicative inverse}

@@

% --------------------------------------------------------------------------
\subsection{Fixpoints\index{Fixpoint} modulo 26}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
%\pagebreak
\hypertarget{Chapter_ElementaryNT_9}{}
\section[Multiplicative order and primitive roots]
           {Multiplicative order and primitive roots\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can have a
    playful experience with primitive roots
    (see learning unit 2.2, pages 10-14/40 and 24-40/40).\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.
}
\label{MultOrdPrimitveRoot}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\hypertarget{Chapter_ElementaryNT_10}{}
\hypertarget{RSABeweis}{}
\section{Proof of the RSA procedure with Euler-Fermat}
\index{RSA}\label{rsabeweis}
Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA!RSA procedure} is the most
common asymmetric cryptography procedure. Developed in 1978 by Ronald Rivest,
Adi Shamir and Leonard Adleman, it can be used both for signatures and for
encryption. Cryptographers always associate this procedure with the
abbreviation ``\textbf{RSA}'' $-$ the following remark is meant with humor to show
that each letter combination can be used with several meanings: In Britain
the ``Royal Society for the encouragement of Arts, Manufactures \& Commerce''
is commonly known as the ``RSA''.
}$^,$\footnote{%
In literature\index{Literature} and in movies\index{Movies} not only 
classic but also modern cryptographic methods have been used 
%(\hyperlink{appendix-movies}
%{see appendix A3: Movies and Literature with Relation to Cryptography}).
(see appendix \ref{s:appendix-movies}).
% hier die items nicht einrücken ! xxxxxxxxxxxxxxx
% \begin{list}{\textbullet}{\leftmargin10pt}
%   \item the movie ``House of Cards'', 1992
%   \item the story ``The Dialogue of the Sisters'' by Dr.~C.~Elsner
% \end{list}\vspace{-\baselineskip}\vspace{-\parskip}
} %end footnote
procedure in the group $\mathbb{Z}_n^*$.

% --------------------------------------------------------------------------
\hypertarget{Chapter_ElementaryNT_10_1}{}
\subsection{Basic idea of public key cryptography}
\label{Label_Chapter_ElementaryNT_10_1}
\index{Cryptography!public key}
